---
output:
  md_document:
    variant: markdown_github
---

# BURGr nibblR 1: Basic data manipulation with Iris

Hello all BURGr lovers! Welcome to the first BURGr bites exercise :smile:. The dataset that we are working with in this exercise is the beloved iris, and our goal is to create a simple, beautiful summary of the data. 

Before we dive into things, let's get an idea of what we are dealing with...

```{r setup, echo=FALSE, }
# http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories#24585750
# http://stackoverflow.com/questions/32128690/knitr-rmd-to-md-and-the-yaml-header
  library(knitr)
  opts_knit$set(root.dir=normalizePath('../'))
  opts_chunk$set(fig.path = "../fig/", dev='png') # corrected path and added dev
  opts_chunk$set(out.format = "md")
```

```{r, data_snippet}
data(iris)
knitr::kable(head(iris))
```

**Part 1:**<a id="part1"></a> The aim is to produce a simple summary of the iris dataset by producing a new dataset from iris that contains the average (i.e. mean ) value of each of the variables (columns) according to species.

Hint: I wonder how one could aggregate this?

```{r, excercise_answer_1}
# work out which column is the "Species" variable
speciesCol = grep("Species", names(iris))
# aggregate iris by species excluding the non-numeric "Species" variable
mean_iris = aggregate(x = iris[, -speciesCol], by = list(iris$Species), FUN = mean, na.rm= T)

# we need to rename all of the columns
names(mean_iris)[1] = "Species"
names(mean_iris)[-1] = paste0(names(mean_iris)[-1], ".mean")
# display pretty table using kable function from knitr
knitr::kable(mean_iris)
```

**Part 2:**<a id="part2"></a> The aim is similar to the aim in [part 1](#part1) but the difference is to also include information about the standard deviation (sd).

Hint: Is this a mind meld or a merge op?

```{r, excercise_answer_2}
# work out which column is the "Species" variable
speciesCol = grep("Species", names(iris))
# aggregate iris by species excluding the non-numeric "Species" variable
sd_iris = aggregate(x = iris[, -speciesCol], by = list(iris$Species), FUN = sd, na.rm= T)

# we need to rename all of the columns
# warning: we are cheating a bit because we know that "Species" is the first variable
# However, most times we will not have this luxury. 
# Question: What would you do then?
names(sd_iris)[1] = "Species"
names(sd_iris)[-1] = paste0(names(sd_iris)[-1], ".sd")

# merge mean and sd data summaries
iris_snapshot = merge(mean_iris, sd_iris, by = "Species")

# similar strategy as above...same caveat (warning)
# Answer: Use grep to avoid hard coding by replacing the 
# number 1 with the results of the grep() search!
speciesCol = grep("Species", names(iris_snapshot))
sortedColNames = c(names(iris_snapshot)[speciesCol], sort(names(iris_snapshot)[-speciesCol]))

# reorganise data by column
iris_snapshot = iris_snapshot[, sortedColNames]

# display pretty table using kable function from knitr
knitr::kable(iris_snapshot)
```

**Part 3:**<a id="part3"></a> In [part 1](#part1) and [part 2](#part2), we were able to create pretty decent summaries of Species data in **iris** according to both the average and standard deviation. 

Hmmm... something is missing? can we make it more "publication-worthy" (sic)... yes we can!

Hint: I sense some **function**al pasting... :alien:

```{r, excercise_answer_3}
# just to mix things up... we will create a new data.frame object
# using existing data

# HTML character code from here: https://www.w3schools.com/charsets/ref_html_entities_4.asp
# needed to add the ability to round down the sd
# makePlusMinus = function(x, y) return(paste(x, "&#177;", y))
makePlusMinus = function(x, y) return(paste(round(x, 2), "&#177;", round(y, 2)))

# create a new data frame out using the "Species" column from existing data
iris_plus = mean_iris["Species"]
# use the custom makePlusMinus to make pretty summaries of the data
iris_plus$Sepal.Length = makePlusMinus(mean_iris$Sepal.Length.mean, sd_iris$Sepal.Length.sd)
iris_plus$Sepal.Width = makePlusMinus(mean_iris$Sepal.Width.mean, sd_iris$Sepal.Width.sd)
iris_plus$Petal.Length = makePlusMinus(mean_iris$Petal.Length.mean, sd_iris$Petal.Length.sd)
iris_plus$Petal.Width = makePlusMinus(mean_iris$Petal.Width.mean, sd_iris$Petal.Width.sd)

# display pretty table using kable function from knitr
knitr::kable(iris_plus)
```

**Part 4:** This bite sized chunk of BURGr goodness is complete, thanks for joining us! The open question is, can you think of other ways to create a simple, pretty **tabular** summary of this dataset?